\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
draft of first two intro paragraphs:
1) talk about why Cox would rather parametric and hazard of hazard ratios. 2) introduce efron and arjas who categorize time and use logistic. then explain that H&M dont categorize time. they select person moments in continuous time

In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ... and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation]. Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances. Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards]. The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR. Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].

>>>>

The semiparametric Cox model has become the default approach to survival analysis even though Cox himself later suggested he would prefer to model the hazard function directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ... and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation]. Indeed, the most relevant quantity in a clinical setting is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances. Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards]. The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR. While stepwise survival curves can be computed with the Cox model, they require a second step to separately estimate the baseline hazard [@breslow1972discussion].

[-@arjas1987logistic] and [-@efron1988logistic] were some of the first authors to pursue approaches that made the fitting of smooth hazard functions much more transparent and intuitive through generalized linear models. They focused on categorizing the time axis (e.g days) and treated each patient-day as a Bernoulli random variable with probability equal to the discrete hazard rate. Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] selected a discrete set of person-time coordinates (`person-moments') in continuous time from all observed follow-up experience constituting the study base to obtain the logistic form likelihood expression. More specifically, all person-moments when the event of interest occurred are selected as the case series, complemented by a randomly sampled base series of person-moments serving as controls.

[-@copenhagen2012needs] has an R package that uses as the slices of person time as the units of analysis. Can keep subdividing them, but in the limit the file becomes infinite. Cite estimation issue from IRLS in Kalbfleish and Prentice.

Just like Efron, we can use the well-understood logistic regression for everything. We are not saying it is better, just an attractive alternative (and a general framework). One that with our package, is well within the reach of more endusers.

We also bring hazard functions out of the closet. Where they tend to be kept by the Cox Model.

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.

Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk. Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach. And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.

### 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
draft of first two intro paragraphs:
1) talk about why Cox would rather parametric and hazard of hazard ratios. 2) introduce efron and arjas who categorize time and use logistic. then explain that H&M dont categorize time. they select person moments in continuous time

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \pkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\end{itemize}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit a
flexible hazard by using a smooth function of time modeled with cubic
B-splines. The model is fit using \code{fitSmoothHazard} with the
familiar formula interface:

\begin{Schunk}
\begin{Sinput}
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#>     data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.403  -0.172  -0.135  -0.085   3.801  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5787.6  on 54535  degrees of freedom
#> AIC: 5798
#> 
#> Number of Fisher Scoring iterations: 8
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\begin{Soutput}
#>  time                
#>     0 0.0e+00 0.0e+00
#>     1 5.4e-05 4.2e-05
\end{Soutput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from
\pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details
section, the usual asymptotic results hold for likelihood ratio tests
built using case-base sampling models. Therefore, we can easily test the
significance of the spline model:

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
#> NULL                               54539       6059             
#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
#> ScrArm              1        8     54535       5788   0.0048 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The null hypothesis is rejected in favor of the spline model. Similarly,
to compare different models (e.g.~time modeled linearly), we could
compute Akaike's Information Criterion (AIC) for each model and compare
them.

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIF for control and screening groups in the ERSPC data. We plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line). We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{Schunk}

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}\\%
%
\email{sahir.bhatnagar@mcgill.ca}%
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}\\%
%
\email{max.turgeon@umanitoba.ca}%
}

\address{%
Jesse Islam\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
%
%
\email{jesse.islam@mail.mcgill.ca}%
}

\address{%
James A. Hanley\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
%
\email{james.hanley@mcgill.ca}%
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}\\%
%
\email{olli.saarela@utoronto.ca}%
}

>>>>

In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked
how he would model a set of censored survival data, to which he
responded: ``I think I would normally want to tackle problems
parametrically \ldots{} and if you want to do things like predict the
outcome for a particular patient, it's much more convenient to do that
parametrically'' \citep{reid1994conversation}. Indeed, the most relevant
quantity in a clinical setting is often the 5- or 10-year risk of
experiencing a certain event given the patient's particular
circumstances. Unfortunately, the most reported metric from a Cox model
is the (potentially time-dependent) hazard ratio (HR), which ignores the
duration of follow-up or is subject to selection bias
\citep{hernan2010hazards}. The covariate adjusted survival curve
overcomes these limitations, and is arguably a more important summary
measure to report than the HR. Stepwise survival curves can be computed
with the Cox model but require a second step to separately estimate the
baseline hazard \citep{breslow1972discussion}.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} for fitting fully parametric hazard models and
covariate adjusted survival curves using the familiar interface of the
\code{glm} function. Using a parametric estimator of the hazard function
leads to a smooth function of time; as a consequence, the survival
function estimates also vary smoothly over time.

Other packages can do this.

the familiar interface of the \code{glm} function. It also provides
tools for variable selection, competing-risk analysis, and dealing with
time-dependent exposures. In addition, we provide functions for
exploratory data analysis and visualizing the estimated quantities.

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using the
  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
  This functionality is provided through the \pkg{glmnet} package
  \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \CRANpkg{mgcv} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \CRANpkg{gbm} package. This feature is still
  experimental.
\end{itemize}

@@

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit
four models that differ in which functional form of time is used: 1)
excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
models are fit using \code{fitSmoothHazard} with the familiar formula
interface:

\begin{Schunk}
\begin{Sinput}
fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(form) {
  fitSmoothHazard(form, data = ERSPC, ratio = 100)
})
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we
leveraged the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fits[["splines"]]) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.409  -0.172  -0.135  -0.084   3.808  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5785.3  on 54535  degrees of freedom
#> AIC: 5795
#> 
#> Number of Fisher Scoring iterations: 9
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0, 14, by = 0.1)

risks <- lapply(fits, function(fit) {
  absoluteRisk(fit, time = new_time, newdata = new_data)
})
\end{Sinput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
\pkg{casebase} on the Cox model CIF. The CIF estimates for the
exponential model in panel (1) overestimate the cumulative incidence
earlier on in time, and underestimate it later on. Based on our earlier
discussion of the population-time plot, this poor fit for the
exponential hazard was expected. We notice a better fit with increasing
complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
(panels 2--4). As noted in Theoretical Details section, the usual
asymptotic results hold for likelihood ratio tests built using case-base
sampling models. Therefore, we can easily test the null hypothesis that
the exponential model is just as good as the larger (in terms of number
of parameters) spline model:

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
#> 1     54538       6052                         
#> 2     54535       5785  3      267   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The null hypothesis is rejected in favor of the spline model. Similarly,
the AIC provides further evidence that the flexible function of time
provides the best fit:

\begin{Schunk}
\begin{Soutput}
#>     Exp. Gompertz  Weibull  Splines 
#>     6056     5821     5807     5795
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time. (1) The time variable is excluded (exponential). (2) Linear function of time (Gompertz). (3) The natural logarithm (Weibull). (4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{Schunk}

In Table \ref{tab:print-erspc-estimates}, we present a side-by-side
comparison of the hazard ratios and confidence intervals estimated from
\code{fitSmoothHazard} and the corresponding parametric model using
\code{survival::survreg}, as well as the Cox model estimate. The hazard
ratio estimates and confidence intervals are similar across all four
models. This reinforces the idea that, under proportional hazards, we do
not need to model the full hazard to obtain reliable estimates of the
hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that
different parametric models can still give rise to qualitatively
different estimates for the CIF.

\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
\centering
\begin{tabular}[t]{>{}lcc}
\toprule
Model & casebase::fitSmoothHazard & survival::survreg\\
\midrule
\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
\end{tabular}
\end{table}

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://sahirbhatnagar.com/}%
%
\\\email{sahir.bhatnagar@mcgill.ca}
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}%
%
\\\email{max.turgeon@umanitoba.ca}
}

\address{%
Jesse Islam\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
%
%
\\\email{jesse.islam@mail.mcgill.ca}
}

\address{%
James A. Hanley\\
McGill University\\%
1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
%
\\\email{james.hanley@mcgill.ca}
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}%
%
\\\email{olli.saarela@utoronto.ca}
}
